---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit README.Rmd file -->

# `MiXcan: Statistical Framework for Cell-type-Aware Transcriptome-Wide Association Studies with Bulk Tissue Data`

## Introduction to **MiXcan**

**Goal:** 

* Constructs cell-type-level prediction models for genetically regulated expression (GReX);

* Predicts cell-type-level GReX in new genotype data; and

* Performs cell-type-aware TWAS.

**Advantages over tissue-level TWAS:**

* Improves GReX prediction accuracy; 

* Boosts the study power, especially for genes that function in minor cell types or have different association directions in different cell types;

* Sheds light on the responsible cell type(s) of  associations. 

**Disadvantages over tissue-level TWAS:**

* Requires prior knowledge on disease-critical cell types and their proportions in tissue;

* Has more model parameters;

* May be less powerful than tissue-level TWAS for genes that have similar disease associations in different cell types or function in major cell types. 


**Input:**

* Prediction model construction: genotype, covariates, and gene expression data (same as in PrediXcan) + cell-type composition estimates (e.g. from existing methods, such as ESTIMATE, CIBERSORT, xCell). 

* Association Analysis: genotype, covariates and phenotype data (same as in PrediXcan).


**Output:**

* Prediction model construction: Cell-type-specific or nonspecific prediction weights for different genes. 

* Association Analysis: Tissue-level association p-values and cell-type-level association summaries including estimates, standard error and p-values.




A full description of the method can be found in our [paper](https://www.medrxiv.org/content/10.1101/2025.11.25.25341027v1).

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(MiXcan2) 
```


## Installation

### Hardware Requirements

The MiXcan package requires only a standard computer with a reasonable RAM/CPU to support the operations. The minimal RAM recommended by authors is 2 GB. Authors used a computer with the following specification:

RAM: 32 GB

CPU: 2.3 GHz 8-core

### Software Requirements

The github package is developed on macOS operating systems. The MiXcan pacakge is developed under an open source software R (version 4.1.2). Different versions of R can be downloaded at https://cran.r-project.org/. 

### Package Installation

With R, users can install the MiXcan package directly from GitHub with [devtools](https://github.com/hadley/devtools):

```R
install.packages("devtools")
devtools::install_github("songxiaoyu/MiXcan2/Package")

````
The typical install time of the package is less than 5 minutes. 

## Example of use

Below demonstrates the MiXcan analysis pipeline on a single peusdo gene. In reality, multiple genes can be analyzed in parallel. Holdout set can be pre-excluded to allow model training on the remaining samples. 

### Data

Create peusdo data for demonstration:

```{r}
# Pseudo data creation
gene_name <- "BRCA1"

set.seed(123)
N <- 100  # number of samples
P <- 10   # number of genetic predictors
Q <- 3    # number of covariates

# Covariate matrix (e.g. age, principal components)
Cov_test <- matrix(rnorm(N * Q), nrow = N, ncol = Q)
colnames(Cov_test) <- c("age", "PC1", "PC2")

# Cell-type fraction estimates (bounded between 0 and 1)
Pi_test <- runif(N, min = 0.1, max = 0.9)

# Genotype matrix with V-prefixed column names
X_test <- matrix(sample(0:2, N * P, replace = TRUE), nrow = N, ncol = P)
colnames(X_test) <- paste0("V", sample(40:100, P))

# SNP annotation matrix
X_rows_test <- data.frame(
  varID    = paste0("chr21_", sample(1e6:5e6, P), "_",
                    sample(c("A","T","C","G"), P, replace = TRUE), "_",
                    sample(c("A","T","C","G"), P, replace = TRUE), "_b38"),
  position = sample(1e6:5e6, P),
  rsid     = paste0("rs", sample(1000:9999, P)),
  ref      = sample(c("A","T","C","G"), P, replace = TRUE),
  eff      = sample(c("A","T","C","G"), P, replace = TRUE)
)

# Expression levels
Y_test <- rnorm(N, mean = 0, sd = 1)
```

### MiXcan analysis pipeline
```{r}
library(doParallel)
library(tidyverse)
library(rlist)
nCores=detectCores()-1; registerDoParallel(nCores) # use parallel computing for speed, but leave 1 core out for other activities. 

```

Step 2: Estimating cell-type-specific (and nonspecific) GReX prediction weights of a gene using the MiXcan function

```{r}
# X_test
#      V70 V53 V59 V79 V63 V76 V80 V68 V49 V88
# [1,]   1   1   0   2   0   1   2   0   2   1
# [2,]   0   2   2   0   2   2   2   2   1   0
# [3,]   0   0   2   2   0   2   1   0   2   2
# [4,]   2   1   2   1   2   0   1   0   2   2
# [5,]   0   0   1   0   2   1   1   1   2   0
# [6,]   2   1   2   0   1   2   2   2   1   0

# X_rows_test
#             xNameMatrix position   rsid ref eff
# 1 chr21_4380166_A_A_b38  4598800 rs2457   G   C
# 2 chr21_2424043_A_C_b38  4171961 rs5424   A   T
# 3 chr21_1669969_T_C_b38  4261569 rs9509   C   G
# 4 chr21_1455554_T_G_b38  2729790 rs7777   C   A
# 5 chr21_4849901_G_A_b38  2055678 rs9293   C   C
# 6 chr21_2125566_T_T_b38  3991202 rs6735   C   A

# Cov_test
#             age         PC1        PC2
# [1,] -0.5604757 -0.7104066  2.1988103
# [2,] -0.2301775  0.2568837  1.3124130
# [3,]  1.5587083 -0.2466919 -0.2651451
# [4,]  0.0705084 -0.3475426  0.5431941
# [5,]  0.1292877 -0.9516186 -0.4143399
# [6,]  1.7150650 -0.0450277 -0.4762469

# Pi_test
# [1] 0.2897838 0.6491923 0.2806547 0.3547957 0.2391871 0.7411437

# Y_test (numeric vector of length N)
# [1]  1.7791027  0.5351380 -0.3719449 -1.0255422 -0.5824017  0.3428884
```


```{r}
set.seed(123)
ensbl <- MiXcan2_ensemble(y = Y_test, x = X_test, cov = Cov_test,
                              pi = Pi_test, xNameMatrix = X_rows_test,
                              yName = gene_name, B = 3, seed = 123)
ensbl$ensemble_intecept
```

Step 3: Extracting the weights and model summaries from the MiXcan output.
Raw SNP weights from every individual ensemble model (across all B iterations). No averaging applied across B model interations and retaining all weights including 0.

```{r}
all_weights_df <- ensbl$all_weights
head(all_weights_df)
```
Overall averaged model performance metrics (e.g. cross-validated R², number of SNPs) Results are summarized across all B models into a single row per gene
```{r}
ensbl_summary_by_type =ensbl$ensemble_summary_by_type
ensbl_summary_by_type
```

Model performance metrics from each ensemble model (across B interations). Reults are prior to averaging, one row per model iteration.

```{r}
ensbl_all_summary =ensbl$all_summary 
ensbl_all_summary
```
